Procurement Dynamics: Analyzing Federal Contract Awards to Microsoft

### 2024-12-03 paper/paper.qmd
Add files via upload

---
title: "Procurement Dynamics: Analyzing Federal Contract Awards to Microsoft and Bell"
subtitle: "Trends and Insights of Microsoft’s Future in Canadian Government from 2020"
author: 
  - Deyi Kong
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/Procurement/](https://github.com/eeeee-cmd/Procurement/)."
date: today
date-format: long
abstract: "This paper analyzes the federal procurement contracts awarded to Microsoft, focusing on trends in award values, contract categories, and the relationships between buyers and Microsoft. The analysis reveals that Microsoft dominates high-value contracts and demonstrates consistent partnerships with key government departments. This paper utilizes a linear model to predict the future contract amount trend for Microsoft. By examining these procurement dynamics, this paper providing insights into the broader implications of tech-industry participation in government procurement."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load libraries
library(tidyverse)
library(sf)
library(dplyr)
library(here)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
library(modelsummary)

@@

The main estimand in this study is the amount of federal procurement contracts awarded to Microsoft. This metric is analyzed across multiple dimensions, including contract categories, durations, and partnerships, to quantify Microsoft's role in federal procurement comprehensively. The findings reveal that Microsoft plays a dominant role in IT infrastructure and modernization, securing the majority of high-value contracts and establishing consistent partnerships with departments such as Employment and Social Development Canada and National Defence. These insights highlight the broader implications of the private tech sector’s influence on public sector procurement, raising questions about competitiveness, dependency, and strategic alignment.

@@

PreparatoryPhase, PhaseDays, and ContractDays variables were created for the purpose of analyzing preparatory phase and performance phase corresponding to each contract. PreparatoryPhase calculates the days from AwardDate to StartDate. ContractDays calculates the days from StartDate to EndDate. PhaseDays take the absolute value of PreparatoryPhase because of the minor effect of either prepare before contract being awarded or prepare after contract being awarded.

@@

In this analysis, several key predictors were identified to evaluate procurement data effectively. The first variable, **Contract**, is a character-type variable that identifies each procurement contract awarded to Microsoft. [@tbl-contract] [SOME MODIFICATION OF CONTRACT CATEGORIES CLEANING PROCESS MAY APPLY]

@@

# Summarize the data and filter buyers with counts >= 25
summary_table <- cleaned_data %>%
  group_by(ContractType) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 50) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

```{r}
#| label: tbl-buyer
#| tbl-cap: The table summarizes top 10 buyer's name who has over 50 contracts with Microsoft in total with contract count, total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 50
summary_table <- cleaned_data %>%
  group_by(Buyer) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 25) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

**Supplier** is a character-type variable includes all the affiliated subsidiary companies of Microsoft. [@tbl-supplier] shows the top 5 buyer's name who has awarded over 50 contracts. [DELETE THIS VARIABLE? OR MAYBE MOVE IT TO APPDX.]

```{r}
#| label: tbl-supplier
#| tbl-cap: The table summarizes the main Microsoft supplier's name who has over 25 contracts with buyers in total with contract count, total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 50
summary_table <- cleaned_data %>%
  group_by(Supplier) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 50) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

# Display the summary table using kableExtra for nicer formatting
summary_table %>%
  kable(col.names = c("Suppllier", "Contract", "Total Amount", "Average")) %>%
  kable_styling(full_width = F, position = "center", bootstrap_options = c("striped", "hover"))

@@

@@

To understand the factors influencing the amount of federal procurement contracts awarded to Microsoft. I construct a linear regression model to explains variations in contract values.

@@

This paper utilizes a linear regression model to examine the relationships between the dependent variable.

$$Amount_i = \beta_0 + \beta_1 ContractType + \beta_2 Buyer + \beta_3 ContractDays + \beta_4 PhaseDays + \epsilon_i$$
As defined in [@sec-data], each predictor variable has been chosen to reflect characteristics that influence the contract amount.

### Model Justification
We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

@@

The model takes several assumptions, including the relationship between the predictors and the outcome is linear, each contract is independent, the variance of the error term is constant, and predictor variables are not highly correlated with each other.

While the model presented offers valuable insights into the dynamics of Microsoft’s federal procurement contracts, there are some limitations including categorization of contract types, generalizability, data quality, potential for endogeneity, and unobserved variables. These limitations will be discussed in [@sec-discussion].

## Model Validation [REMOVE?]

R-squared, adjusted R-squared

# load data and model
lm_model_m <- readRDS(here::here("models/lm_model_microsoft.rds"))
# lm_model_b <- readRDS(here::here("models/lm_model_bell.rds"))

# model_list <- list("Microsoft Model" = lm_model_m, "Bell Model" = lm_model_b)
modelsummary(model_list, 
             output = "kableExtra", 
             statistic = "std.error", 
             coef_map = c("ContractType" = "Contract Type", 
                          "Buyer" = "Buyer", 
                          "ContractDays" = "ContractDays",
                          "PhaseDays" = "PhaseDays")) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

# Summary of the Random Forest model
summary(rf_model)
# Print the model summary
print(rf_model)
# Error rates
rf_model$err.rate  # Error rates for each class (classification)
# Access the OOB error rate
rf_model$oob.times  # Times each observation was used in the model
rf_model$oob.err    # Out-of-bag error estimate

@@

# Create a new variable for "four-month periods"
data <- cleaned_data %>%
  mutate(
    Year = year(AwardDate),
    Month = month(AwardDate),
    Period = case_when(
      Month %in% 1:4 ~ "Jan-Apr",
      Month %in% 5:8 ~ "May-Aug",
      Month %in% 9:12 ~ "Sep-Dec"
    ),
    YearPeriod = paste(Year, Period, sep = "-")  # Combine Year and Period
  )

# Count the number of contracts in each four-month period
contract_counts <- data %>%
  group_by(YearPeriod) %>%
  summarise(Count = n())

# Heatmap plot
ggplot(contract_counts, aes(x = YearPeriod, y = Count, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "red") +
  labs(x = "Year-Period", y = "Frequency (Contract Count)", title = "Heatmap of Contracts by Four-Month Periods") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0))  # Keep y-axis labels readable
```

@@

@@

@@

# Appendix {#sec-appx}

## Methodology

@@

### Data Sources and Structure

@@

### Data Integration and Challenges:

@@

### Solicitation Methods and Contract Types:

>>>>

---
title: "Procurement Dynamics: Analyzing Federal Contract Awards to Microsoft"
subtitle: "Trends of Microsoft’s Future in Canadian Government from 2020"
author: 
  - Deyi Kong
thanks: "Code and data are available at: [https://github.com/eeeee-cmd/Procurement/](https://github.com/eeeee-cmd/Procurement/)."
date: today
date-format: long
abstract: "This paper analyzes the federal procurement contracts awarded to Microsoft, focusing on trends in award values, contract categories, and the relationships between buyers and Microsoft. The analysis reveals that Microsoft dominates high-value contracts and demonstrates consistent partnerships with key government departments. This paper utilizes a linear model and random forest to predict the future contract amount trend for Microsoft. By examining these procurement dynamics, this paper providing insights into the broader implications of tech-industry participation in government procurement."
format: pdf
toc: true
toc-depth: 3
number-sections: true
bibliography: references.bib
---

@@

# load libraries
library(tidyverse)
library(sf)
library(dplyr)
library(here)
library(arrow)
library(ggplot2)
library(janitor)
library(purrr)
library(knitr)
library(kableExtra)
library(modelsummary)
library(ggcorrplot)
library(gridExtra)
library(rstanarm)

@@

The main estimand in this study is the amount of federal procurement contracts awarded to Microsoft. This metric is analyzed across multiple dimensions, including contract categories, duration, and partnerships, to quantify Microsoft's role in federal procurement comprehensively. The findings reveal that Microsoft plays a dominant role in IT infrastructure and modernization, securing the majority of high-value contracts and establishing consistent partnerships with departments such as Employment and Social Development Canada and National Defence. These insights highlight the broader implications of the private tech sector’s influence on public sector procurement, raising questions about competitiveness, dependency, and strategic alignment.

@@

PreparatoryPhase, PhaseDays, and ContractDays variables were created for the purpose of analyzing preparatory phase and performance phase corresponding to each contract. PreparatoryPhase calculates the days from AwardDate to StartDate. ContractDays calculates the days from StartDate to EndDate. PhaseDays take the absolute value of PreparatoryPhase because of the minor effect of either prepare before contract being awarded or prepare after contract being awarded. Last but not least, I created a new variable called ContractType, which categorizes the contracts based on keywords.

@@

In this analysis, several key predictors were identified to evaluate procurement data effectively. The first variable, **ContractType**, is a character-type variable that identifies each procurement contract awarded to Microsoft into Categories. [@tbl-contract]

@@

# Summarize the data and filter buyers with counts >= 25
summary_table <- cleaned_data %>%
  group_by(ContractType) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 25) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

```{r}
#| label: tbl-buyer
#| tbl-cap: The table summarizes top 10 buyer's name who has over 25 contracts with Microsoft in total with contract count, total dollar amount, and average dollar amount per contract. 
#| echo: false
#| eval: true
#| warning: false
#| message: false

# Summarize the data and filter buyers with counts >= 25
summary_table <- cleaned_data %>%
  group_by(Buyer) %>%
  summarise(
    Count = n(), 
    TotalAmount = sum(Amount),  # Calculate the total contract value
    AvgAmount = mean(Amount)    # Calculate the average contract value
  ) %>%
  filter(Count > 15) %>%
  arrange(desc(Count))  # Arrange in descending order based on Count

@@

@@

```{r}
#| label: fig-cor
#| fig-cap: This correlation matrix visualizes the relationships between key numerical predictor variables and the outcome variable, Amount. The figure shows correlations include a weak negative correlations between PreparatoryPhase and ContractDays; and weak positive correlations between predictors and Amount.
#| echo: false
#| eval: true
#| warning: false
#| message: false

# filter to numerical predictor variables
numerical_data <- cleaned_data %>%
  select(Amount, PreparatoryPhase, ContractDays)

correlation_matrix <- cor(numerical_data, use = "complete.obs")

# create correlation matrix
ggcorrplot(correlation_matrix, lab = TRUE) + 
  theme(
    axis.text.x = element_text(size = 10),  # Adjust x-axis text size
    axis.text.y = element_text(size = 10) # Adjust y-axis text size
  )
```

```{r}
#| label: fig-box
#| fig-cap: The Boxplots shows the relationship of ContractDays and PhaseDays with ContractType. 
#| fig-subcap: ["Contract Duration in each Contract Type","Phase Duration in each Contract Type"]
#| echo: false
#| warning: false
#| layout-nrow: 2

# Clean and transform the data
data <- cleaned_data %>%
  mutate(
    ContractType = gsub("Related", "", ContractType), # Remove "Related"
    ContractType = str_wrap(gsub("/", " or ", ContractType), width = 10) # Wrap long labels
  )

# First graph: ContractType vs. ContractDays
ggplot(data, aes(x = ContractType, y = ContractDays)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Contract Type", y = "Contract Duration") +
  theme(axis.text.x = element_text(size = 6))

# Second graph: ContractType vs. PhaseDays
ggplot(data, aes(x = ContractType, y = PhaseDays)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Contract Type", y = "Phase Duration") +
  theme(axis.text.x = element_text(size = 6))

@@

@@

This paper utilizes a Bayesian model to examine the factors influencing the dollar amount of federal procurement contracts awarded to Microsoft. Define $Amount_i$ as the dollar amount of the federal procurement contract awarded to Microsoft and $\mu_i$ is the expected value of the contract amount modeled as a linear combination of the predictors. The model for predicting $Amount_i$ is specified as follows:
\begin{align*} 
Amount_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma)\\
\mu_i &= \alpha + \beta_1 \cdot \text{ContractType}_i + \beta_2 \cdot \text{Buyer}_i + \beta_3 \cdot \text{ContractDays}_i + \beta_4 \cdot \text{PhaseDays}_i \\
\alpha &\sim \text{Normal}(0, 2) \\
\beta_1, \beta_2, \beta_3, \beta_4 &\sim \text{Normal}(0, 2) \\
\sigma &\sim \text{Exponential}(1)
\end{align*}
Where:\
- $\alpha$ is the intercept term.\
- $\beta_1, \beta_2, \beta_3, \beta_4$ are the coefficients associated with the predictors:\
  - $ContractType_i$: A categorical variable representing the contract type.\
  - $Buyer_i$: A categorical variable representing the buyer of the contract.\
  - $ContractDays_i$: A numeric variable representing the duration of the contract is active.\
  - $PhaseDays_i$: A numeric variable representing the days between AwardDate and StartDate.

The model is fit using the `rstanarm` package of @rstanarm in R [@citeR], with the specified priors guiding the model while allowing it to adapt to the data.

### Model Justification
This paper expect that the factors - such as ContractType, Buyer, ContractDays and PhaseDays - will influence the outcome of contract amount. In particular, ...

In this paper, I compared three different modeling approaches to examine the factors influencing the federal procurement contract amount: linear regression, random forest, and Bayesian regression. Each of these models offers different advantages and trade-offs. More detailed justification about model selection will be discussed in [@sec-discussion]. The Bayesian model is selected due to better performance and flexibility.

Bayesian approach can explicitly model uncertainty by incorporating prior knowledge and provides a full posterior distribution of the model parameters. This is useful when the data is sparse and prior knowledge about the contract types, buyers, or duration is available. The linear regression structure was selected due to its simplicity and interpretability. The numeric dependent variable, $Amount_i$, gives a chance of linear regression for modeling the relationship between the contract amount and its predictors.

By including weakly informative priors for the coefficients and residual standard deviation, the model remains flexible and interpretable, while still preventing extreme parameter estimates. The use of an exponential prior for $\sigma$ allows the model to adapt a scale of residuals in the data, providing estimates of the variance. The normal prior for the coefficients $\beta_1, \beta_2, \beta_3, \beta_4$ with mean 0 and standard deviation 2 shows the expectation that the coefficients should not be too larg but flexible enough to adapt the data.

This Bayesian linear regression model balances complexity with interpretability, and its flexibility allows us to draw meaningful observations from the data while quantifying uncertainty around the estimates, especially when the sample size is not large or when the predictors have a complex relationship with the outcome. [TOO GENERAL, CHANGE]

@@

The model takes several assumptions, including linear relationship between predictors, normality of the residuals, independence of observations, homoscedasticity, and reasonable chosen priors.

While the model presents some valuable observations of Microsoft’s federal procurement contracts, there are some model limitations including prior function selection, data quality, potential for endogeneity, generalizability, and unobserved variables. These limitations will be discussed more detailed in [@sec-discussion].

@@

# Create a new variable for "four-month periods"
data <- cleaned_data %>%
  mutate(
    Period = case_when(
      Month %in% 1:4 ~ "Jan-Apr",
      Month %in% 5:8 ~ "May-Aug",
      Month %in% 9:12 ~ "Sep-Dec"
      )
  )

# Count the number of contracts in each four-month period
contract_counts <- data %>%
  group_by(Period) %>%
  summarise(Count = n())

# Heatmap plot
ggplot(contract_counts, aes(x = Period, y = Count, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(x = "Year-Period", y = "Frequency (Contract Count)", title = "Heatmap of Contracts by Four-Month Periods") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0))  # Keep y-axis labels readable
```

@@

## Model
R-squared, adjusted R-squared

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

bayesian_model <-
  readRDS(file = here::here("models/bayesian_model.rds"))
```

```{r}
#| label: tbl-modelresults
#| echo: false
#| eval: true
#| tbl-cap: "Explanatory models of contract amount based on ContractType, Buyer, ContractDays, PhaseDays and Month"
#| warning: false
#| message: false

# Generate the model summary

bayesian_model_full <- readRDS(here::here("models/bayesian_model_full.rds"))
bayesian_model <- readRDS(here::here("models/bayesian_model.rds"))

model_list <- list("Bayesian Model_Full" = bayesian_model_full, "Bayesian Model" = bayesian_model)

modelsummary(
  model_list,
  output = "kableExtra", 
  coef_map = c("ContractType" = "ContractType", "Buyer" = "Buyer", 
               "ContractDays" = "ContractDays", "PhaseDays" = "PhaseDays")
) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

@@

## Model Selection

@@

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Bayesian Model

### Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(bayesian_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(bayesian_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set3") +
  theme(
    legend.position = "none", 
    axis.text.y = element_text(size = 7)
    ) +
  coord_flip()

### Diagnostics

@fig-trace-1 is a trace plot. It shows... This suggests...

@fig-trace-2 is a trace plot. It shows... This suggests...

@fig-rhat is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot - Contract Type","Trace plot - Buyers"]
#| layout-nrow: 2

# Extract the parameter names
param_names <- colnames(as.matrix(bayesian_model))

# Use grep to find all parameters that contain "ContractType"
contract_type_params <- grep("ContractType", param_names, value = TRUE)

# Use grep to find all parameters that contain "Buyer"
buyer_params <- grep("Buyer", param_names, value = TRUE)

# # Group 1: Buyers containing the word "the"
# buyers_with_the <- grep("the", buyer_params, value = TRUE, ignore.case = TRUE)
# 
# # Group 2: Buyers containing the word "the"
# buyers_with_and <- grep("and", buyer_params, value = TRUE, ignore.case = TRUE)
# buyers_with_and <- buyers_with_and[!grepl("the", buyers_with_and, ignore.case = TRUE)]
# 
# # Group 3: Buyers containing the word "the"
# buyers_with_of <- grep("of", buyer_params, value = TRUE, ignore.case = TRUE)
# buyers_with_of <- buyers_with_of[!grepl("the|and", buyers_with_of, ignore.case = TRUE)]
# 
# # Group 4: Buyers containing the word "Canada"
# buyers_with_canada <- grep("Canada", buyer_params, value = TRUE, ignore.case = TRUE)
# buyers_with_canada <- buyers_with_canada[!grepl("the|and|of", buyers_with_canada, ignore.case = TRUE)]
# 
# # Group 5: Buyers that do not contain "the" or "Canada"
# other_buyers <- grep("Canada|the|and|of", buyer_params, value = TRUE, invert = TRUE, ignore.case = TRUE)

# Plot trace for ContractType
plot(bayesian_model, "trace", pars = contract_type_params) +
  theme_minimal() +
  theme(
    legend.position = "bottom",                # Place legend at the bottom
    legend.text = element_text(size = 6),      # Adjust legend label size
    legend.title = element_text(size = 8),     # Adjust legend title size
    axis.text.x = element_text(size = 6),      # Adjust x-axis text size
    axis.text.y = element_text(size = 6),      # Adjust y-axis text size
    strip.text = element_text(size = 5)
    )

# Plot trace for ContractType
plot(bayesian_model, "trace", pars = buyer_params) +
  theme_minimal() +
  theme(
    legend.position = "bottom",                # Place legend at the bottom
    legend.text = element_text(size = 6),      # Adjust legend label size
    legend.title = element_text(size = 8),     # Adjust legend title size
    axis.text.x = element_text(size = 4),      # Adjust x-axis text size
    axis.text.y = element_text(size = 5),      # Adjust y-axis text size
    strip.text = element_text(size = 5)
    )

# # Plot trace for buyers' name containing "the"
# plot(bayesian_model, "trace", pars = buyers_with_the)
# 
# # Plot trace for buyers' name containing "and"
# plot(bayesian_model, "trace", pars = buyers_with_and)
# 
# # Plot trace for buyers' name containing Canada
# plot(bayesian_model, "trace", pars = buyers_with_canada)
# 
# # Plot trace for buyers' name containing "of"
# plot(bayesian_model, "trace", pars = buyers_with_of)
# 
# # Plot trace for Other buyers
# plot(bayesian_model, "trace", pars = other_buyers)
```
```{r}
#| label: fig-rhat
#| echo: false
#| eval: true
#| fig-cap: "Rhat plot for bayesian model"
#| message: false
#| warning: false
#| fig-subcap: ["Rhat plot - Full model","Rhat plot - Reduced model"]
#| layout-ncol: 2

plot(bayesian_model_full, "rhat")
plot(bayesian_model, "rhat")
```
## Linear Regression Model

# load data and model
lm_model_full <- readRDS(here::here("models/lm_model_full.rds"))
lm_model <- readRDS(here::here("models/lm_model.rds"))
model_list <- list("Linear Model_Full" = lm_model_full, "Linear Model" = lm_model)

modelsummary(
  model_list,
  output = "kableExtra", 
  statistic = "std.error", 
  coef_map = c("ContractType" = "ContractType", "Buyer" = "Buyer", 
               "ContractDays" = "ContractDays", "PhaseDays" = "PhaseDays")
  ) %>%
  kable_styling(latex_options = c("striped", "scale_down"))
```

## Random Forest Model

# Summary of the Random Forest model
summary(rf_model, output = "kableExtra")

# # Error rates
# rf_model$err.rate  # Error rates for each class (classification)
# rf_model$oob.err    # Out-of-bag error estimate

# Methodology

@@

## Data Sources and Structure

@@

## Data Integration and Challenges:

@@

## Solicitation Methods and Contract Types:
